
title: "Reproducible Research Course - Peer Assessment 1"
author: "Created by Fang"
date: "Saturday, June 13, 2015"
output: html_document
keep_md: true


## Step 1: Setting Global Options

```{r setoptions, echo=TRUE, cache = TRUE} # make code visible for reviewers
library(knitr)
opts_chunk$set(echo = FALSE)
```

## Step 2: Set Working Directory to where the repo was cloned 

```{r, echo=TRUE}
setwd("C:/Users/fangl_000/Documents/GitHub/RepData_PeerAssessment1")
getwd()
```

## Step 3: unzip and load the data

```{r, echo=TRUE}
unzip("activity.zip")
mydata <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
head(mydata)
dim(mydata)
summary(mydata)
```

## Step 4: Clean and preprocess the data, create a data without missing value for steps
```{r}
mydata$month <- as.numeric(format(mydata$date, "%m"))
mydata$day <- as.numeric(format(mydata$date, "%d"))
mydata$year <- as.numeric(format(mydata$date, "%y"))
head(mydata)
```

## Step 5: What is mean total number of steps taken per day?
### For this part of the assignment, you can ignore the missing values in the dataset.
  1. Calculate the total number of steps taken per day
  2. Make a histogram of the total number of steps taken each day
  3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
library(ggplot2)

totSteps <-aggregate(na.omit(mydata)$steps, by=list(Group.Date = na.omit(mydata)$date), FUN=sum)$x
print(totSteps)
mean_tot <- mean(totSteps)
median_tot <- median(totSteps)
mean_tot
median_tot

ggplot(mydata, na.rm=TRUE, aes(date, steps)) + geom_bar(stat = "identity", colour = "green", fill = "green", width = 0.6) + facet_grid(. ~ month, scales = "free") + labs(title = "Total Number of Steps Per-Day", x = "Date", y = "Total number of steps")
```

The mean is `r mean_tot` and the median is `r median_tot`.


## Step 6: What is the average daily activity pattern?
   1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
   2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
   
```{r}   
avgSteps <- aggregate(na.omit(mydata)$steps, list(interval = as.numeric(as.character(na.omit(mydata)$interval))), FUN = "mean")
names(avgSteps)[2] <- "AverageSteps"
dim(avgSteps)
head(avgSteps)
ggplot(avgSteps, na.rm = TRUE, aes(interval, AverageSteps)) + geom_line(color = "purple", size = 0.6) + labs(title = "Time Series Plot of 5-minute Intervals", x = "5-minute intervals", y = "Average Number of Steps Taken All Days")
```

```{r}
avgSteps[avgSteps$AverageSteps == max(avgSteps$AverageSteps), ]
```

## Step 7: Imputing missing values
   1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
   
```{r}
   tot_na <- sum(is.na(mydata))
```
Total number of missing value is `r tot_na` 

   2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean   
      /median for that day, or the mean for that 5-minute interval, etc.
      try1: use the mean for the 5 minute interval to fill in missing value for the interval
      try2: use the median for the 5 minute interval to fill in missing value for the interval, but per observation, a lot of rows with median value 0, so it may not    
      make sense to use median
   3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}    
      MedSteps <- aggregate(na.omit(mydata)$steps, list(interval = as.numeric(as.character(na.omit(mydata)$interval))), FUN = "median")
      names(MedSteps)[2] <- "MedianSteps"
#print(MedSteps)
```

```{r}
     Try1Data <- mydata
 for (i in 1:nrow(Try1Data)) {
         if (is.na(Try1Data$steps[i])) {
            Try1Data$steps[i] <- avgSteps[which(Try1Data$interval[i] == avgSteps$interval), ]$AverageSteps
            }
      }
dim(Try1Data)
head(Try1Data)
sum(is.na(Try1Data))

# using ifelse
#Try2Data <- mydata
#ifelse(is.na(Try2Data$steps), Try2Data$steps <-
#for (i in 1:nrow(Try2Data)) {
 #   if (is.na(Try2Data$steps[i])) {
#        Try2Data$steps[i] <- MedSteps[which(Try2Data$interval[i] == MedSteps$interval), ]$MedianSteps
 #   }
#}

#dim(Try2Data)
#head(Try2Data)
#sum(is.na(Try2Data))
```
   4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
totSteps1 <-aggregate(Try1Data$steps, by=list(Group.Date = Try1Data$date), FUN=sum)$x
print(totSteps1)
mean_tot1 <- mean(totSteps1)
median_tot1 <- median(totSteps1)
mean_tot1
median_tot1

ggplot(Try1Data, na.rm=TRUE, aes(date, steps)) + geom_bar(stat = "identity", colour = "red", fill = "red", width = 0.6) + facet_grid(. ~ month, scales = "free") + labs(title = "Total Number of Steps Per-Day", x = "Date", y = "Total number of steps")
```

Using the mean step for that interveral to fill missing value, the new mean is `r mean_tot1` (the original is `r mean_tot`);
The new median is `r median_tot1` (the original is `r median_tot`);
For this data, the impact of imputing missing data on the estimate of the total daily number of steps has no big change as mean stays the same, new median is a little bigger than the original one.


## Step 8: Are there differences in activity patterns between weekdays and weekends?
   1. For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
      Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or 
      weekend day.
      
```{r}
head(Try1Data)
Try1Data$day <- weekdays(as.Date(Try1Data$date))
head(Try1Data)
levels(Try1Data$day)
levels(Try1Data$day) <- list(weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
                             weekend = c("Saturday", "Sunday"))
levels(Try1Data$day)
table(Try1Data$day)
```
   2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}   
avgSteps <- aggregate(Try1Data$steps, list(interval = as.numeric(as.character(Try1Data$interval)), 
                                           day = Try1Data$day), FUN = "mean")
names(avgSteps)[3] <- "AverageSteps"

dim(avgSteps)
head(avgSteps)


library(lattice)

xyplot(avgSteps$AverageSteps ~ avgSteps$interval | avgSteps$day, 
       layout = c(1, 2), type = "l", 
       xlab = "5 Minutes Interval", ylab = "Average Number of Steps")

#ggplot(avgSteps, aes(interval ~ AverageSteps)|day) + geom_line(color = "purple", size = 0.6) + labs(title = "Time Series Plot of 5-minute Intervals", layout = c(1, 2), x = "5-minute intervals", y = "Average Number of Steps Taken All Days")
```



